---
title: "Reproducible Reseach Assiignment 1"
output: html_document
author: "Abhay"
date: "2026-03-31"
---

##loading and processing the data

```{r echo=TRUE}
 data <-read.csv("activity.csv")
 data$date <- as.Date(data$date)
```

##mean total no. of steps
```{r echo=TRUE}
steps_per_day <- aggregate(steps ~ date, data = data, sum, na.rm = TRUE)
head(steps_per_day)
```
##histogram making
```{r histogram,echo=TRUE}
hist(steps_per_day$steps,
     main = "Total Steps per Day",
     xlab = "Steps",
     col = "blue")
```
##mean & median calculation
```{r mean_&_median, echo=TRUE}
mean_steps <- mean(steps_per_day$steps)
median_steps <- median(steps_per_day$steps)

mean_steps
median_steps
```

##average daily activity pattern

```{r avg_daily_pattern, echo=TRUE}
interval_avg <- aggregate(steps ~ interval, data = data, mean, na.rm = TRUE)
```

##time series plot
```{r echo=TRUE}
plot(interval_avg$interval, interval_avg$steps,
     type = "l",
     xlab = "Interval",
     ylab = "Average Steps")
```
##max interval

{r max_intervalecho=TRUE}interval_avg[which.max(interval_avg$steps), ]

##report the total number of missing values

```{r echo=TRUE}
sum(is.na(data$steps))
```

##mean per interval imputing

```{r echo=TRUE}
filled_data <- data
for(i in 1:nrow(filled_data)){
  if(is.na(filled_data$steps[i])){
    interval_val <- filled_data$interval[i]
    mean_val <- interval_avg$steps[interval_avg$interval == interval_val]
    filled_data$steps[i] <- mean_val
  }
}
```
##histogram after imputing

```{r echo=TRUE}
steps_per_day_filled <- aggregate(steps ~ date, data = filled_data, sum)

hist(steps_per_day_filled$steps,
     main = "Total Steps per Day (Filled)",
     xlab = "Steps",
     col = "green")
```

##mean, median mode after imputing

mean_filled <- mean(steps_per_day_filled$steps)
```{r echo=TRUE}
steps_per_day_filled <- aggregate(steps ~ date, data = filled_data, sum)
median_filled <- median(steps_per_day_filled$steps)
mean_filled <- mean(steps_per_day_filled$steps)
  
mean_filled
median_filled
```
## are there diff between weekdays and weekends
```{r echo=TRUE}
filled_data$day <- ifelse(weekdays(filled_data$date) %in% c("Saturday", "Sunday"),
                          "weekend", "weekday")

filled_data$day <- as.factor(filled_data$day)
```

## avg steps by interval and day types
library(lattice)

```{r average_steps_by_interval, echo=TRUE}
avg_day_type <- aggregate(steps ~ interval + day, data = filled_data, mean)

xyplot(steps ~ interval | day,
       data = avg_day_type,
       type = "l",
       layout = c(1,2),
       xlab = "Interval",
       ylab = "Steps")
```
##end if sccript